Error Reporting Parsers: a Monad Transformer Approach

### 2013-06-21 article.tex
Parser article: split error reporting into two separate sections

\subsection{Reporting position}
To report error position, we'll pre-process the input
string to calculate the line and column number of each character.  Then, instead 
of our parsers operating on a list of characters, they'll operate on a list of 
Tokens, where a Token is a character tupled up with a line number and column 
number, which we'll express with a simple typedef and some accessor functions:
\EZY{I don't think this is how most real parser libraries do it.  Why not do it
with some extra state on the parser monad; it's yet another application of the
transformer approach?}
\Matt{Good point, I thought about doing it this way but decided not to because
the article is already long and this would add a lot, and the other approach 
seems to work fine and is simple.
But I'm not opposed to it at all.  Do you think I should switch it?}
\EZY{I think that it is a worthwhile change to make, if it doesn't add a lot
of extra stuff to the article.  I think it could be done that way, but if
you don't think so I'm happy to let this lie.}
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a 
list of characters into a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
\EZY{In contrast, this section doesn't seem very enlightening, so maybe
it could be glossed over or relegated to an appendix.}
\Matt{I don't quite understand -- I'm missing the context.
What does "in constrast" refer to?"}
\EZY{Sorry, I was referring to the terseness of the first section. So there
you give very little explanation, while here you spend a lot of time explaining
what is going on.}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with a similar one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, 
              Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}

Now our token parsers are:
\begin{verbatim}
whitespace = many1 $ satisfy (flip elem " \n\t\r\f" . chr)
comment = pure (:) <*> character ';' <*> many0 chars
  where chars = not1 $ character '\n'

munch p = many0 (whitespace <+> comment) *> p

ocurly = munch $ character '{'
ccurly = munch $ character '}'
oparen = munch $ character '('
cparen = munch $ character ')'
symbol = munch $ many1 char
  where char = fmap chr $ satisfy (f . chr)
        f = flip elem (['a' .. 'z'] ++ ['A' .. 'Z'])
\end{verbatim}

The only differences were the replacement of \verb+literal+ with \verb+character+ 
and a couple uses of the \verb+chr+ accessor.  The \verb+munch+ combinator didn't 
have to change at all (although its inferred type did in fact change).

\subsection{Error messages}
Now, we modify the remaining parsers to report accurate and useful errors.
Here are the error messages that we'll use to report each error:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

@@

Now let's start with \verb+application+ as our first modified, error-reporting 
parser:
\begin{verbatim}
application =
    openparen                             >>= \open ->
    commit (eAppOper, open) form          >>= \op ->
    many0 form                            >>= \args ->
    commit (eAppClose, open) closeparen   >>
    return (AApp op args)
\end{verbatim}
This parser works by running the open-parenthesis parser, capturing its value 
in \verb+open+, and continuing with the parsing; if either \verb+form+ or 
close-parenthesis fails, an appropriate error is generated with a message and 
the location of the open parenthesis.  However, if the open 
parenthesis is not successfully parsed, it's a failure but not an error.  
Examples:
\begin{verbatim}
*Main> runParser application (countLineCol "oops")
Right Nothing

*Main> runParser application (countLineCol "(a b c)")
Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],[]))

*Main> runParser application (countLineCol "\n()")
Left ("application: missing operator",('(',2,1))

*Main> runParser application (countLineCol "(a b")
Left ("application: missing close parenthesis",('(',1,1))
\end{verbatim}

Although the previous version of the \verb+application+ parser was implemented 
with just applicative combinators, this version can't be:  \verb+open+ 
is used to build the following parsers, making this monadic.

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    opencurly                            >>= \open ->
    check (== "define") symbol            *>
    pure ADefine                         <*>
    commit (eDefSym, open) symbol        <*>
    commit (eDefForm, open) form         <*
    commit (eDefClose, open) closecurly

lambda =
    opencurly                                >>= \open ->
    check (== "lambda") symbol               >>
    commit (eLamParam, open) opencurly       >>= \p_open ->
    many0 symbol                             >>= \params ->
    (if distinct params 
        then return ()
        else throwError (eLamDupe, p_open))  >>
    commit (eLamPClose, p_open) closecurly   >>
    commit (eLamBody, open) (many1 form)     >>= \bodies ->
    commit (eLamClose, open) closecurly      >>
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}

@@

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be parsed.  We extend the \verb+special+ 
parser with a third alternative, which throws an error if it can parse
an open curly brace:
\begin{verbatim}
special = define   <+> 
          lambda   <+> 
          (ocurly >>= \o -> throwError (esName, o))
\end{verbatim}

@@

When we can no longer parse any forms, instead of failing, we'd like to report 
an error if there's any input left, and otherwise succeed.  Our ending parser
must first munch any leading whitespace or comments, as usual, and then check 
whether there are any tokens left.  If there are, it should report an error at 
the location of the first remaining token:
\begin{verbatim}
endCheck = 
    many0 (whitespace <+> comment) *>
    get >>= \xs -> case xs of
                        (t:_) -> throwError (ewUnp, t)
                        []    -> pure ()
\end{verbatim}
The \verb+woof+ parser uses \verb+endCheck+ after consuming all possible forms:
\begin{verbatim}
woof = many0 form <* endCheck
\end{verbatim}

\subsection{Error messages:  examples}
While correct input is still parsed successfully:
\begin{verbatim}
*Main> let p = runParser woof . countLineCol

*Main> p "{define fgh {lambda {f x} (f x)}}"
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") 
                                           [ASymbol "x"]])],
       []))
\end{verbatim}

Incorrect input results in an error.  For each of the items in our error spec, 
we can show that our parser correctly reports an error: 
\begin{verbatim}
*Main> p "()"
Left ("application: missing operator",('(',1,1))

*Main> p "(a b"
Left ("application: missing close parenthesis",('(',1,1))

*Main> p "{define (a b c)}"
Left ("define: missing symbol",('{',1,1))

*Main> p "{define a}"
Left ("define: missing form",('{',1,1))

*Main> p "{define a b"
Left ("define: missing close curly",('{',1,1))

*Main> p "{lambda (a b c)}"
Left ("lambda: missing parameter list",('{',1,1))

*Main> p "{lambda {a b a} (a b c)}"
Left ("lambda: duplicate parameter names",('{',1,9))

*Main> p "{lambda {a b (c d)}"
Left ("lambda: missing parameter list close curly",('{',1,9))

*Main> p "{lambda {a b}}"
Left ("lambda: missing body form",('{',1,1))

*Main> p "{lambda {a b} (c d)"
Left ("lambda: missing close curly",('{',1,1))

*Main> p "{defin x y}"
Left ("special form: unable to parse",('{',1,1))

*Main> p "a,b"
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}

>>>>

\subsection{Error messages}
We must modify the remaining parsers to report meaningful error messages; let's
define the messages that will be reported in each case:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

@@

Now let's start with \verb+application+ as our first modified, error-reporting 
parser:
\begin{verbatim}
application =
    openparen                     >>
    commit eAppOper form          >>= \op ->
    many0 form                    >>= \args ->
    commit eAppClose closeparen   >>
    return (AApp op args)
\end{verbatim}

This parser works by first running the open-parenthesis parser; then, if either 
\verb+form+ or close-parenthesis fails, an appropriate error is generated with 
a message describing the problem.  However, if the open 
parenthesis is not successfully parsed, it's a failure but not an error.  
Examples:
\begin{verbatim}
*Main> runParser application "oops"
Right Nothing

*Main> runParser application "(a b c)"
Right (Just (AApp (ASymbol "a") [ASymbol "b",ASymbol "c"],""))

*Main> runParser application "\n()"
Left "application: missing operator"

*Main> runParser application "(a b"
Left "application: missing close parenthesis"
\end{verbatim}

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    check (== "define") symbol    *>
    pure ADefine                 <*>
    commit eDefSym symbol        <*>
    commit eDefForm form         <*
    commit eDefClose closecurly

lambda =
    check (== "lambda") symbol       >>
    commit eLamParam opencurly       >>
    many0 symbol                     >>= \params ->
    (if distinct params 
        then return ()
        else throwError eLamDupe)    >>
    commit eLamPClose closecurly     >>
    commit eLamBody (many1 form)     >>= \bodies ->
    return (ALambda params bodies)   <*
    commit eLamClose closecurly  
  where
    distinct names = length names == length (nub names)
\end{verbatim}

@@

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be parsed.  We extend the \verb+special+ 
parser to report an error if it matches an open brace but not define or lambda:
\begin{verbatim}
special = 
    opencurly  *>  commit eSpecial (define <|> lambda)
\end{verbatim}

@@

When we can no longer parse any forms, instead of failing, we'd like to report 
an error if there's any input left, and otherwise succeed.  Our ending parser
must first munch any leading whitespace or comments, as usual, and then check 
whether there are any characters left; if there are, it should report an error:
\begin{verbatim}
endCheck = 
    many0 (whitespace <|> comment) *>
    switch item
\end{verbatim}
The \verb+woof+ parser uses \verb+endCheck+ after consuming all possible forms:
\begin{verbatim}
woof = many0 form <* commit eWoof endCheck
\end{verbatim}

\subsection{Error messages:  examples}
While correct input is still parsed successfully:
\begin{verbatim}
*Main> runParser woof "{define fgh {lambda {f x} (f x)}}"
Right (Just ([ADefine "fgh" (ALambda ["f","x"] 
                                     [AApp (ASymbol "f") 
                                           [ASymbol "x"]])],
       []))
\end{verbatim}

Incorrect input results in an error.  For each of the items in our error spec, 
we can show that our parser correctly reports an error: 
\begin{verbatim}
*Main> runParser woof "(a b"
Left "application: missing close parenthesis"

*Main> runParser woof "{define (a b c)}"
Left "define: missing symbol"
\end{verbatim}

And so on.

\section{Example 3: reporting position}

... new code ...
